GENEX: Transcript Report

### 2020-09-08 code/02-Transcript_Master.Rmd
MetaAnalysisCode

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(readr)
library(biomaRt)
library(knitr)
library(kableExtra)
library(tidyverse)
library(grid)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(yaml)
knitr::opts_chunk$set(
  #root.dir = '/home/jgockley/AD_TargetRank/',
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)

@@

Gene <- "GENEX"
#Gene <- "CTHRC1"

@@

### Log2 Fold Change of Transcripts By Brain Region Case V. Control
Red indicates over expressed and Blue indicates under expressed in cases versus Controls \newline
AD = CDR >= 1 & BRAAK >= 4 & CERAD >= 2 \newline
CONTROL = CDR <= 0.5 & BRAAK <= 3 & CERAD <= 1 \newline

@@

### Transcript Expression Level Boxplot Case V. Control
Red indicates over expressed and Blue indicates under expressed in cases versus Controls \newline
AD = CDR >= 1 & BRAAK >= 4 & CERAD >= 2 - Red \newline
CONTROL = CDR <= 0.5 & BRAAK <= 3 & CERAD <= 1 - Blue \newline
Other - Green

@@

}else{
  if(length(LIST) > 4 ){
    PLOTS <- paste0(LIST, collapse=',')
    eval(parse(text= paste0( 'grid.arrange(', PLOTS , ', ncol=4, nrow=2 )' )))
  }else{
    PLOTS <- paste0(LIST, collapse=',')
    eval(parse(text= paste0( 'grid.arrange(', PLOTS , ', ncol=4, nrow=1 )' )))
  }
}

#grid.arrange(DLPFC, TCX, CBE, PHG, IFG, FP, STG,ncol=4, nrow=2)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(readr)
library(biomaRt)
library(knitr)
library(kableExtra)
library(tidyverse)
library(grid)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(yaml)
library(MASS)
knitr::opts_chunk$set(
  #root.dir = '/home/jgockley/AD_TargetRank/',
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)

@@

Gene <- "GENEX"
#Gene <- "CTHRC1"
#Gene <- "GFAP"
#Gene <- 'C4B'

@@

### Log2 Fold Change of Transcripts By Brain Region Case V. Control
Red indicates over expressed and Blue indicates under expressed in cases versus Controls \newline
AD = CDR >= 1 & Braak >= 4 & CERAD >= 2 \newline
CONTROL = CDR <= 0.5 & Braak <= 3 & CERAD <= 1 \newline

@@

### Log2 Fold Change of Transcripts By Cognitive Assesment
Red indicates over expressed and Blue indicates under expressed in cases versus Controls \newline
ROSMAP: Cogdx = 4 Versus CogDx = 1
\begin{itemize}
\item[] Cogdx = 4: AD: Alzheimer’s dementia and NO other cause of CI (NINCDS PROB AD)
\item[] Cogdx = 1: NCI: No cognitive impairment (No impaired domains)
\end{itemize}

MSBB: CDR = 5 Versus CDR = 0
\begin{itemize}
\item[] CDR = 5: Severe cognitive decline
\item[] CDR = 0: No cognitive decline
\end{itemize}

```{r Cognition, eval = T, include=F}
DLPFC <- Log_Plot( Rosmap_DE, 'DLPFC', 'cogdx4-cogdx1', 'cogdx', Rosmap_Adj )

PHG <- Log_Plot( MSBB_DE, 'PHG', 'CDR', 'SourceDiagnosis', MSBB_Adj  )
IFG <- Log_Plot( MSBB_DE, 'IFG', 'CDR', 'SourceDiagnosis', MSBB_Adj  )
FP <- Log_Plot( MSBB_DE, 'FP', 'CDR', 'SourceDiagnosis', MSBB_Adj  )
STG <- Log_Plot( MSBB_DE, 'STG', 'CDR', 'SourceDiagnosis', MSBB_Adj  )

LIST <- NULL
for( test in c('DLPFC','PHG','IFG','FP','STG') ){
  if( eval(parse(text= paste0( 'is.null(', test, ')' ))) == TRUE ){
  }else{
    LIST <- c(LIST, test)
  }
}
```

```{r PlotLogCog, cache = FALSE, fig.height = 5, fig.width = 14}
N<-length(LIST) 
if(length(LIST) == 0 ){

```{r pagebreak2, results='asis', eval=is_latex_output()}
cat('\\pagebreak')
```

### Neuropathology Associations
All associations are logistic ordinal regression models preformed on Braak Score with P-Values correct to the total number of transcripts belonging to the user input gene list expressed in the respective tissue. \newline

```{r WrangleNeuroPath, eval = T, include=F }
MSBB_N   <- read.table( file = '~/AD_TargetRank/Transcript_Profiles/Temp_Files/MSBB_NeuroPath.tsv', header = T, sep = '\t', stringsAsFactors = F )
row.names(MSBB_N) <- MSBB_N$sampleIdentifier 
ROSMAP_N <- read.table( file = '~/AD_TargetRank/Transcript_Profiles/Temp_Files/Rosmap_NeuroPath.tsv', header = T, sep = '\t', stringsAsFactors = F )
row.names(ROSMAP_N) <- ROSMAP_N$Sampleid 
Mayo_N   <- read.table( file = '~/AD_TargetRank/Transcript_Profiles/Temp_Files/Mayo_NeuroPath.tsv', header = T, sep = '\t', stringsAsFactors = F )
row.names(Mayo_N) <- Mayo_N$SampleID

MSBB_Braak <- MSBB_N[ , c( "sampleIdentifier", "Tissue1", "bbscore") ]
colnames(MSBB_Braak) <- c( "Sample", 'Tissue', "Braak" )
ROSMAP_N$Tissue <- 'DLPFC'
ROSMAP_Braak <- ROSMAP_N[ , c("Sampleid", "Tissue", "braaksc") ] 
colnames(ROSMAP_Braak) <- c( "Sample", 'Tissue', "Braak" )
Mayo_Braak <- Mayo_N[ ,c("SampleID","Tissue","Braak")]
colnames(Mayo_Braak) <- c( "Sample", 'Tissue', "Braak" )

Braak <- as.data.frame(rbind(Mayo_Braak,ROSMAP_Braak,MSBB_Braak))
Braak$Sample <- as.character( Braak$Sample )
Braak$Tissue <- as.character( Braak$Tissue )

Results <- data.frame( matrix( NA, 1 ,10 ) )
colnames( Results ) <- c( 'Transcript', 'Gene', 'Symbol', 'Tissue', 'Coeff', 
                          '95_Conf_L', '95_Conf_U', 'Z_Score', 'P_Val', 'FDR_PVal'
                        )

# Plot Function
Braak_Plot <- function( data, tiss, gene ){
  #'@data data frame eg Results
  #'@tiss the tissue to use eg DLPFC
  #'@gene the gene to use eg gene

  eval(parse(text=paste0( 'pDat <- data[ data$Symbol %in% gene & data$Tissue == \'', tiss ,'\', ]' )))

  if( dim(pDat)[1] > 0 ){

    if( (tiss %in% c('DLPFC','TCX','PHG')) == T){
      plot <- ggplot(  pDat   , aes(x=Coeff, y=Transcript , colour=Direction)) +
        scale_y_discrete() + 
        scale_color_manual(values=c("#000000", "#7F0000" , "#000099"), breaks =c("NONE", "UP" , "DOWN")) +
        geom_errorbar(aes(xmin=`95_Conf_L`, xmax=`95_Conf_U`), width=.1) +
        geom_line() +
        geom_point() +
        #annotation_custom(grob) 
        ggtitle(paste0(tiss,': ',gene)) + theme(legend.position = "none") + theme(axis.title.y = element_blank()) +
        theme(axis.text.y = element_text(face="bold", size=6, angle=45)) +
        geom_vline(xintercept = 0, linetype="dotted", color = "black", size=1.5)
    }else{
       plot <- ggplot(  pDat   , aes( x=Coeff, y=Transcript, colour=Direction )) +
        scale_y_discrete() + 
        scale_color_manual(values=c("#000000", "#7F0000" , "#000099"), breaks =c("NONE", "UP" , "DOWN")) +
        geom_errorbar(aes(xmin=`95_Conf_L`, xmax=`95_Conf_U`), width=.1) +
        geom_line() +
        geom_point() +
        #annotation_custom(grob) 
        ggtitle(paste0(tiss,': ',gene)) + theme(legend.position = "none") + 
        theme(axis.title.y = element_blank(), axis.text.y = element_blank()) +
        geom_vline(xintercept = 0, linetype="dotted", color = "black", size=1.5)
    } 
  }else{
    plot <- NULL
  }
  return(plot)
}

#RunAnalysis  
for( TI in c( 'DLPFC', 'TCX', 'CBE', 'PHG', 'IFG', 'STG', 'FP' ) ){
  eval(parse(text= paste0( 'Braak_temp <- Braak[ Braak$Tissue == \'', TI, '\', ]' )))
  Braak_temp$Braak <- as.factor( Braak_temp$Braak )

  if( TI == 'DLPFC' ){
    #table(row.names(Rosmap_EXP) %in% Trans$ensembl_transcript_id)
    EXP_temp <- Rosmap_EXP[,Braak_temp$Sample]
  }else{
    if( (TI %in% c('TCX','CBE')) == T  ){
      EXP_temp <- Mayo_EXP[,Braak_temp$Sample]
    }else{
      if( (TI %in% c( 'PHG', 'IFG', 'STG', 'FP')) == T  ){
        EXP_temp <- MSBB_EXP[,Braak_temp$Sample]
      }else{
        warning("Improper Tissue Type")
      }
    }
  }

  row.names(Trans) <- Trans$ensembl_transcript_id

  if( any(Trans[ Trans$hgnc_symbol == Gene,]$ensembl_transcript_id %in% row.names(EXP_temp)) ){
    Pull <- Trans[ Trans$hgnc_symbol == Gene,]$ensembl_transcript_id[ Trans[ Trans$hgnc_symbol == Gene,]$ensembl_transcript_id %in% row.names(EXP_temp) ]
    EXP_temp <- EXP_temp[ Pull , ]

    for( i in 1:dim(EXP_temp)[1]){

      Transcript <- row.names(EXP_temp)[i]
      Ge <- Trans[Transcript,]$hgnc_symbol
      ENSG <- Trans[Transcript,]$ensembl_gene_id

      foo <- cbind( Braak_temp[,1:3], 
                    t(EXP_temp[ row.names(EXP_temp)[i], row.names(Braak_temp) ])
                  )
      foo$Braak <- as.factor(foo$Braak)
      colnames(foo)[4] <- "Ge"
      m <- polr( Braak ~ Ge, data = foo, Hess=TRUE)

      summary(m)
      ctable <- coef(summary(m))
      p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
      ctable <- cbind(ctable, "p value" = p)
      ci <- confint(m)

      Adj_P <- p.adjust( ctable['Ge','p value'], method = 'fdr', n = dim(EXP_temp)[1] )

      Entry <- c( Transcript, ENSG, Ge, TI, ctable['Ge','Value'], 
                  as.numeric(ci[1]), as.numeric(ci[2]), ctable['Ge', "t value"] , ctable['Ge','p value'],
                  Adj_P 
                )
      Results <- rbind( Results[1:dim(Results)[1], ], Entry)
    }
  }
  else{
  }
  Results <- Results[ complete.cases(Results), ]
  if( dim(Results[ complete.cases(Results), ])[1] > 0){
    Results$Direction <- 'NONE'
    Results$Direction[Results$FDR_PVal <= 0.05 & Results$Z_Score < 0 ] = 'DOWN'
    Results$Direction[Results$FDR_PVal <= 0.05 & Results$Z_Score > 0] = 'UP'

    Results$Coeff <- as.numeric(as.character( Results$Coeff ))
    Results$`95_Conf_L` <- as.numeric(as.character( Results$`95_Conf_L` ))
    Results$`95_Conf_U` <- as.numeric(as.character( Results$`95_Conf_U` ))
    Results$Z_Score <- as.numeric(as.character( Results$Z_Score ))
    Results$P_Val  <- as.numeric(as.character( Results$P_Val ))
    Results$FDR_PVal <- as.numeric(as.character( Results$FDR_PVal ))

    eval(parse(text= paste0( '\'', TI, '\' <- Braak_Plot( Results, \'', TI, '\', Gene )' )))
  }else{
  }
}

```{r PlotLogBrak, cache = FALSE, fig.height = 5, fig.width = 14}
N<-length(LIST) 
if(length(LIST) == 0 ){

```{r pagebreak3, results='asis', eval=is_latex_output()}
cat('\\pagebreak')
```

### Transcript Expression Level Boxplot Case V. Control
AD = CDR >= 1 & Braak >= 4 & CERAD >= 2 - Red \newline
CONTROL = CDR <= 0.5 & Braak <= 3 & CERAD <= 1 - Blue \newline
Other - Green

@@

}else{
  if(length(LIST) > 4 ){
    PLOTS <- paste0(LIST, collapse=',')
    eval(parse(text= paste0( 'grid.arrange(', PLOTS , ', ncol=4, nrow=2 )' )))
  }else{
    PLOTS <- paste0(LIST, collapse=',')
    eval(parse(text= paste0( 'grid.arrange(', PLOTS , ', ncol=4, nrow=1 )' )))
  }
}
```

```{r pagebreak4, results='asis', eval=is_latex_output()}
cat('\\pagebreak')
```

### Gene Model
Only expressed trascripts i.e. those in the analysis above are used in the gene model. \newline

```{r ExonPlot, eval = T, include=F }

library(Gviz)
library(rtracklayer)
library(trackViewer)
library(org.Hs.eg.db)
library( TxDb.Hsapiens.UCSC.hg38.knownGene )
library("biomaRt")

gene_id <- Gene
filters <- 'hgnc_symbol'
attrs <- c(filters, "ensembl_gene_id", "chromosome_name", "start_position", "end_position", 'strand')

host <- 'ensembl.org'
organism <- 'hsa'
ensembl <- biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host = host)

ds <- biomaRt::listDatasets(ensembl)[, "dataset"]
ds <- grep(paste0("^", organism), ds, value = TRUE)
if (length(ds) == 0) {
  stop(paste("Mart not found for:", organism))
} else if (length(ds) > 1) {
  message("Found several marts")
  sapply(ds, function(d) message(paste(which(ds == d), d, sep = ": ")))
  n <- readline(paste0("Choose mart (1-", length(ds), ") : "))
  ds <- ds[as.integer(n)]
}
ensembl <- biomaRt::useDataset(ds, mart = ensembl)

trans <- getBM(filters = 'hgnc_symbol',
               attributes = attrs,
               values = Gene,
               mart = ensembl )

if( trans$strand < 0 ){
  Strand <- '-'
}else{
  Strand <- '+'
}

# txdb <- makeTxDbFromEnsembl("Homo sapiens", server="useastdb.ensembl.org")
TxDB_ENS <- makeTxDbFromEnsembl("Homo sapiens", server="useastdb.ensembl.org")

#entrezID <- get(Gene, org.Hs.egSYMBOL2EG)
EID <- Trans[ Trans$hgnc_symbol == Gene, ]$ensembl_gene_id[1]
#geneTrack(EID,TxDB_ENS)[[1]]
theTrack <- geneTrack(EID,TxDB_ENS)[[1]]

gr <- GRanges( paste0('chr',trans$chromosome_name ), IRanges( trans$start_position, trans$end_position), strand=Strand)
trs <- geneModelFromTxdb(TxDb.Hsapiens.UCSC.hg38.knownGene,
  org.Hs.eg.db,
  gr=gr)

viewerStyle <- trackViewerStyle()
setTrackViewerStyleParam(viewerStyle, "margin", c(.1, .12, .02, .02))

TIDs <- Trans[ Trans$hgnc_symbol == Gene, ]$ensembl_transcript_id
TIDs <- TIDs[ !duplicated(TIDs) ]
Trs <- trs[ names(trs)[ do.call( rbind, strsplit( names(trs), '[.]' ))[,1]  %in% TIDs ] ]

track <- trackList( Trs)
for( i in 1:length(track) ){
  setTrackStyleParam( track[[i]], "ylabgp", list(cex=.8) )
}

if( any( grepl('MHC', gr@seqnames@values )) ){
  PLOT <- 'NO'
}else{
   PLOT <- 'YES'
}
```

```{r PlotExon, cache = FALSE, fig.height = 8, fig.width = 14}
library(trackViewer)
if( PLOT == 'YES' ){
  viewTracks(track, 
    gr=gr, viewerStyle=viewerStyle, 
    autoOptimizeStyle=TRUE)
}else{
  print("Gene model has some transcripts assigned to alternate chromosomes a more detailed look is Required")
}
